 

### 2023-09-11 markdown/acknowledgments.Rmd

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

>>>>

Beside my faithful feline companion, Apollo, who took care of my place in Austria while I worked on this project in Stockholm, I extend my deep gratitude to all the members of Ujjwal Neogi's research group. Their consistent guidance and support throughout this project have been truly invaluable, and I am profoundly appreciative of their expertise and assistance.

Furthermore, I wish to express my gratitude to Gerhard Dürnberger, my internal supervisor, for his unwavering support and feedback on my master thesis, along with his precious time devoted to our meetings.

### 2023-09-11 markdown/article_doc.Rmd

---
title: " "
output:
  bookdown::pdf_document2:
    toc: false
geometry: "left = 2cm, right = 1.5cm, top = 1.5cm, bottom = 1.5cm"
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \let\origfigure\figure
  - \let\endorigfigure\endfigure
  - \renewenvironment{figure}[1][2] {
      \expandafter\origfigure\expandafter[H]
    } {
    \endorigfigure
    }
  - \usepackage{sectsty}
  - \usepackage{paralist}
  - \usepackage{setspace}\spacing{1.5}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{dcolumn}
  - \usepackage{natbib}\bibliographystyle{agsm}
  - \usepackage[nottoc, numbib]{tocbibind}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multicol,blindtext}

@@

\begin{centering}
```{r uni_logo, echo=F, out.width="20%", fig.align='center'}
knitr::include_graphics("KI_logo.png")
```
This project was performed at the systems virology lab at the Karolinska Institutet in Stockholm.
\end{centering}

@@

```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.align='center', fig.cap= "Flowchart presenting the processing by ProteoScanR. Gold: techniques which involve explanatory data analysis. Yellow: Adjustable features. Blue: Inputs"}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

```{r ui-numbers-default, echo=F, out.width="100%", fig.align='center', fig.cap="Screenshot of proteomics workbench. Summary barplot showing the number of peptide spectrum matches, peptides and proteins"}
knitr::include_graphics("screenshots/numbers_default.png")
```

@@

```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.align='center', fig.cap="Screenshot of statistics module in portrait mode"}
knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
```

>>>>

---
title: " "
output:
  bookdown::word_document2:
    toc: false
geometry: "left = 2cm, right = 1.5cm, top = 1.5cm, bottom = 1.5cm"
fontsize: 11pt
header-includes:
  - \usepackage{float}
  - \let\origfigure\figure
  - \let\endorigfigure\endfigure
  - \renewenvironment{figure}[1][2] {
      \expandafter\origfigure\expandafter[H]
    } {
    \endorigfigure
    }
  - \usepackage{sectsty}
  - \usepackage{paralist}
  - \usepackage{setspace}\spacing{1.5}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{dcolumn}
  - \usepackage{natbib}\bibliographystyle{agsm}
  - \usepackage[nottoc, numbib]{tocbibind}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multicol,blindtext}

@@

\begin{centering}
```{r uni_logo, echo=F, out.width="20%"}
knitr::include_graphics("KI_logo.png")
```
This project was performed at the systems virology lab at the Karolinska Institutet in Stockholm.
\end{centering}

@@

```{r data-processing-pipeline-flowchart-vertical, echo=FALSE, fig.height=27, fig.width=17, out.width="75%", fig.cap= "Flowchart presenting the processing by ProteoScanR. Gold: techniques which involve explanatory data analysis. Yellow: Adjustable features. Blue: Inputs"}
library(Gmisc, quietly = TRUE)
library(glue)
library(htmlTable)
library(grid)
library(magrittr)

@@

```{r ui-numbers-default, echo=F, out.width="100%", fig.cap="Screenshot of proteomics workbench. Summary barplot showing the number of peptide spectrum matches, peptides and proteins"}
knitr::include_graphics("screenshots/numbers_default.png")
```

@@

```{r ui-stat-HC-HIV-noMets, echo=F, out.width="80%", fig.cap="Screenshot of statistics module in portrait mode"}
knitr::include_graphics("screenshots/stat_HC_HIV_noMets.png")
```

### 2023-09-11 markdown/discussion.Rmd

Unfortunately, a substantial number of individual proteins in the data set (totaling `r n_missingness`) have missing values. On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels (see figure \ref@(fig:missingness)). 
This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can sometimes lead to artificially evenly distributed values. Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
To improve future experiments, it is recommended to increase the number of proteins in the carrier proteome. By expanding the repertoire of proteins included in the carrier proteome, it can enhance the coverage and depth of analysis, providing a more comprehensive understanding of the biological system under investigation. This can potentially lead to more robust and accurate results in downstream analyses.

@@

As observed in the dendrogram (see figure \ref@(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data. The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels. This can potentially lead to a decrease in the statistical power of tests aimed at detecting differential expression between sample types. 
Ideally, clustering approaches should result in individual clusters that capture distinct sample types, enabling the testing of differential expression within each cluster. This stratification of samples within clusters allows for the identification of unique expression patterns associated with specific biological variations. The dendrogram, as demonstrated above, is an example of hierarchical clustering, which provides a visual representation of the clustering results. In this context, users should have the flexibility to select the desired number of clusters that best explain the underlying biological variation in their dataset. This allows for a more targeted and precise analysis of differential expression within each cluster, facilitating the identification of biologically meaningful patterns and insights.
To begin with the k-means clustering approach, it is crucial to determine the optimal number of clusters before conducting the analysis. This can be achieved using various approaches, one of which is the gap statistic method \citep{Tibshirani2001}.

@@

The gap statistic (see figure \ref@(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution. The null reference distribution represents a scenario where no clear clustering patterns exist in the data. However as seen in the gap statistic a high number of clusters will be difficult to distinguish from each other and explore the biological meaning such as differentiation in the example macrophage dataset.
Another approach used to determine the optimal number of clusters is the silhouette method.

@@

The silhouette method (see figure \ref@(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters. 
If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k. This suggests that the data points may not be well-separated into distinct clusters. As suggested by the silhouette method in this case a lower number of clusters to differentiate cell types is recommended.

@@

As seen in the principal components analysis (see figure \ref@(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters. The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation. One of the other factors involves the large number of imputed values due the high rate of missing values in the protein expression set. K-NN is not able to impute biological differences and therefore averages the particular missing value by its k neighbors.

>>>>

Unfortunately, a substantial number of individual proteins in the data set (totaling `r n_missingness`) have missing values. On average, this amounts to approximately `r round(mean_missingness, 2)*100`% of the proteins across all runs and channels (see figure \@ref(fig:missingness)). 
This situation leads to the need for missing value imputation using techniques such as k-nearest neighbor (K-NN), which can sometimes lead to artificially evenly distributed values. Additionally, it is important to note that this imputation process can introduce potential loss of power in subsequent statistical analysis.
To improve future experiments, it is recommended to increase the number of proteins in the carrier proteome. By expanding the repertoire of proteins included in the carrier proteome, it can enhance the coverage and depth of analysis, providing a more comprehensive understanding of the biological system under investigation. This can potentially lead to more robust and accurate results in downstream analyses.

@@

As observed in the dendrogram (see figure \@ref(fig:cluster-analysis-dendrogram)), it is evident that sample type alone does not account for the variation observed in the data. The branching patterns in the dendrogram suggest that other factors might be significant for contributing to the averaging of protein expression levels. This can potentially lead to a decrease in the statistical power of tests aimed at detecting differential expression between sample types. 
Ideally, clustering approaches should result in individual clusters that capture distinct sample types, enabling the testing of differential expression within each cluster. This stratification of samples within clusters allows for the identification of unique expression patterns associated with specific biological variations. The dendrogram, as demonstrated above, is an example of hierarchical clustering, which provides a visual representation of the clustering results. In this context, users should have the flexibility to select the desired number of clusters that best explain the underlying biological variation in their dataset. This allows for a more targeted and precise analysis of differential expression within each cluster, facilitating the identification of biologically meaningful patterns and insights.
To begin with the k-means clustering approach, it is crucial to determine the optimal number of clusters before conducting the analysis. This can be achieved using various approaches, one of which is the gap statistic method \citep{Tibshirani2001}.

@@

The gap statistic (see figure \@ref(fig:cluster-analysis-gap-stat)) compares the total intracluster variation across various values of k (number of clusters) to the expected variation under a null reference distribution. The null reference distribution represents a scenario where no clear clustering patterns exist in the data. However as seen in the gap statistic a high number of clusters will be difficult to distinguish from each other and explore the biological meaning such as differentiation in the example macrophage dataset.
Another approach used to determine the optimal number of clusters is the silhouette method.

@@

The silhouette method (see figure \@ref(fig:cluster-analysis-silh)) calculates silhouette coefficients for each data point, which quantify how closely the point resembles its own cluster compared to other clusters. 
If the overall distances between clusters are small, indicating overlapping or unclear separation, it is suggested to take the minimum k. This suggests that the data points may not be well-separated into distinct clusters. As suggested by the silhouette method in this case a lower number of clusters to differentiate cell types is recommended.

@@

As seen in the principal components analysis (see figure \@ref(fig:cluster-analysis-dim-red)), clusters are densely packed, and the plot shows no clear differentiation between the `r optimal_clusters` suggested clusters. The dense packing and poor resolution indicates that other factors may be influencing the lack of distinct separation. One of the other factors involves the large number of imputed values due the high rate of missing values in the protein expression set. K-NN is not able to impute biological differences and therefore averages the particular missing value by its k neighbors.

### 2023-09-11 markdown/introduction.Rmd

## From the Central Dogma of Molecular Biology to Omics and Systems Biology
Before the initial exploration of molecular biology, the processes of life were roughly inferred and not well understood. Among other principles the central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins. This framework consists of two major processes: transcription, which converts DNA into RNA, and translation, which converts RNA into functional proteins. Between these processes, molecular-level regulation occurs through chemical adaptations. This regulation is crucial for biological organisms to efficiently produce their genetic products, known as proteins. Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
Over the past six decades, researchers have explored various domains of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts. Omics research has emerged as an unbiased approach to study biomolecules and their interplay within biological systems comprehensively. Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules. By analyzing multiple layers of omics data, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks using computational methods.
Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks. Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure. Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}

@@

Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their (m/z) consecutively. Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently. The detector transforms the impact (TOF) or oscillation (Orbitrap) energy of the ion into an electric signal, called spectrum which can be processed computationally.

>>>>

## From the Central Dogma of Molecular Biology to Omics and Systems Biology
Before the initial exploration of molecular biology, the processes of life were roughly inferred and not well understood. Among other principles the central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins. This framework consists of two major processes: transcription, which converts DNA into RNA, and translation, which converts RNA into functional proteins. Between these processes, molecular-level regulation occurs through chemical adaptations. This regulation is crucial for biological organisms to efficiently produce their genetic products, known as proteins. Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb2017}.
Over the past six decades, researchers have explored various domains of this principle, collectively referred to as "omics," to comprehensively analyze biomolecules in diverse contexts. Omics research has emerged as an unbiased approach to study biomolecules and their interplay within biological systems comprehensively. Omics encompasses various fields, including genomics, transcriptomics, proteomics, and metabolomics, with each field focusing on a specific class of biomolecules. By analyzing multiple layers of omics data, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks using computational methods.
Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks. Biological systems are imagined as complex, multi-layer networks, with the ensemble of proteins, known as the proteome, playing a critical role within this interactive structure. Proteins have various functions, such as providing structural integrity, catalyzing chemical reactions, and regulating cellular functions \citep{Karahalil2016}.

@@

Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their m/z consecutively. Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently. The detector transforms the impact (TOF) or oscillation (Orbitrap) energy of the ion into an electric signal, called spectrum which can be processed computationally.
